Clustering exploration for `r params$sample_id`

### 2024-11-06 analyses/cell-type-wilms-tumor-06/notebook_template/03_clustering_exploration.Rmd
Fix subdiagnosis bug by adding unique, and fix several typos while we're here

```{r}
subdiagnosis <- readr::read_tsv(
  file.path("..", "..", "..", "data", "current", params$scpca_project_id, "single_cell_metadata.tsv"),
  show_col_types = FALSE
) |>
  dplyr::filter(scpca_sample_id == params$sample_id) |>
  dplyr::pull(subdiagnosis)
```

@@

- in the case each cluster do have a set of specific marker genes, we could expect each of the cluster to have at least a different phenotype,
- if marker genes are mostly shared between clusters, we might have overclustered,
- if no marker genes are found, the quality of cells in the clusters might be impaired (high mitochondrial content? ribosomal content?)

@@

For a Seurat object `object`, the function `MSigDB_score` will calculate a score using `AddModuleScore()` function for a MSigDB gene set `gs_name` in the category `category`

@@

- `category` is a MSigDB collection (https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp).
values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`

- `gs_name` is the name of a MSigDB gene set, e.g. `"HALLMARK_P53_PATHWAY"`

@@

- `category` is the MSigDB category or collection to be used
values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`

@@

Here, we open the table of marker genes marker-sets/CellType_metadata.csv.
Note: we do not expect to have a clear and nice pattern of expression for all of the following markers in every tumor.
This is just ti get a few idea.

@@

#### DROSHA target genes

@@

#### DICER1 target genes

@@

We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense.
DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method.
Please check the preprint from Chistoph Hafemeister: https://www.biorxiv.org/content/10.1101/2023.03.28.534443v1
and tool described here: https://github.com/cancerbits/DElegate

@@

We used three gene sets from MSiGDB:

- the hallmark gene sets  are coherently expressed signatures derived by aggregating many MSigDB gene sets to represent well-defined biological states or processes.
- the C3 : regulatory target gene sets  based on gene target predictions for microRNA seed sequences and predicted transcription factor binding sites.
- the C8 : cell type signature gene sets  curated from cluster markers identified in single-cell sequencing studies of human tissue.

We used enricher function from clusterProfiler to perform enrichment analysis.

@@

#### Hallmarks MSiGDB gene sets

@@

EMT signature should be enriched in stroma cluster.
E2F/proliferation should be enriched in blastema cluster.
MYC(N), TP53 must be enriched in blastema cluster.

#### C8 MSiGDB cell type signature gene sets

@@

The MSigDB C8 gene set is quite relevant for kidney and nephroblastoma annotations.
Epithelial (cancer and normal) cells should be enriched in mature/adulte kidney pathways while blastema cancer cells will show enrichment of fetal kidney development pathway / cap mesenchyme.

@@

Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia [[1](https://www.ncbi.nlm.nih.gov/books/NBK373356/), [2](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/)].
Because of their histological similarity to fetal kidneys, Wilms tumors are thought to arise from developmental derangements in embryonic renal progenitors.

@@

    + Annotations determined by SingleR, an automated reference-based method [Looney et al. 2019](https://www.nature.com/articles/s41590-018-0276-y).

    + Annotations determined by CellAssign, an automated marker-gene based method [Zhang et al. 2019](https://www.nature.com/articles/s41592-019-0529-1).

@@

#### Annotation from SingleR (no cancer or kidney specific dataset)

@@

#### Annotation from cellassign (no cancer or kidney specific dataset)

@@

- `fetal_kidney_predicted.compartment` is one of the 4 main compartments composing a fetal kidney.
We expect immune and endothelial cells to be healthy (non-canerous) cells identified easily with high confidency.
We expect stroma and fetal nephron compartment to contain both normal and malignant cells.

@@

Here we check wether one (or more) `Seurat` cluster do correspond to one label.
This would allow us to be more confident in the annotation of immune or endothelial cells for example.

@@

#### Comparison of the  fetal reference with `singleR` and `cellassign`

>>>>

```{r}
subdiagnosis <- readr::read_tsv(
  file.path("..", "..", "..", "data", "current", params$scpca_project_id, "single_cell_metadata.tsv"),
  show_col_types = FALSE
) |>
  dplyr::filter(scpca_sample_id == params$sample_id) |>
  dplyr::pull(subdiagnosis) |>
  unique()
```

@@

- in the case each cluster do have a set of specific marker genes, we could expect each of the cluster to have at least a different phenotype,
- if marker genes are mostly shared between clusters, we might have over-clustered,
- if no marker genes are found, the quality of cells in the clusters might be impaired (high mitochondrial content? ribosomal content?)

@@

For a Seurat object `object`, the function `MSigDB_score` will calculate a score using `AddModuleScore()` function for a `MSigDB` gene set `gs_name` in the category `category`

@@

- `category` is a `MSigDB` collection (https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp).
values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`

- `gs_name` is the name of a `MSigDB` gene set, e.g. `"HALLMARK_P53_PATHWAY"`

@@

- `category` is the `MSigDB` category or collection to be used
values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`

@@

Here, we open the table of marker genes `marker-sets/CellType_metadata.csv`.
Note: we do not expect to have a clear and nice pattern of expression for all of the following markers in every tumor.
This is just ti get a few idea.

@@

#### `DROSHA` target genes

@@

#### `DICER1` target genes

@@

We run `DElegate::FindAllMarkers2` to find markers of the different clusters and manually check if they do make sense.
`DElegate::FindAllMarkers2` is an improved version of `Seurat::FindAllMarkers` based on pseudobulk differential expression method.
Please check the preprint from Hafemeister: https://www.biorxiv.org/content/10.1101/2023.03.28.534443v1
and tool described here: https://github.com/cancerbits/DElegate

@@

We used three gene sets from `MSigDB`:

- the hallmark gene sets  are coherently expressed signatures derived by aggregating many `MSigDB` gene sets to represent well-defined biological states or processes.
- the C3 : regulatory target gene sets  based on gene target predictions for microRNA seed sequences and predicted transcription factor binding sites.
- the C8 : cell type signature gene sets  curated from cluster markers identified in single-cell sequencing studies of human tissue.

We used enricher function from `clusterProfiler` to perform enrichment analysis.

@@

#### Hallmarks `MSigDB` gene sets

@@

`EMT` signature should be enriched in stroma cluster.
`E2F`/proliferation should be enriched in blastema cluster.
`MYC(N)`, `TP53` must be enriched in blastema cluster.

#### C8 `MSigDB` cell type signature gene sets

@@

The `MSigDB` C8 gene set is quite relevant for kidney and nephroblastoma annotations.
Epithelial (cancer and normal) cells should be enriched in mature/adult kidney pathways while blastema cancer cells will show enrichment of fetal kidney development pathway / cap mesenchyme.

@@

Wilms tumors can contain up to three histologies that resemble fetal kidney: blastema, stroma, and epithelia. [[1](https://www.ncbi.nlm.nih.gov/books/NBK373356/), [2](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9915828/)].
Because of their histological similarity to fetal kidneys, Wilms tumors are thought to arise from developmental derangements in embryonic renal progenitors.

@@

    + Annotations determined by `SingleR`, an automated reference-based method [Looney et al. 2019](https://www.nature.com/articles/s41590-018-0276-y).

    + Annotations determined by `CellAssign`, an automated marker-gene based method [Zhang et al. 2019](https://www.nature.com/articles/s41592-019-0529-1).

@@

#### Annotation from `SingleR` (no cancer or kidney specific dataset)

@@

#### Annotation from `CellAssign` (no cancer or kidney specific dataset)

@@

- `fetal_kidney_predicted.compartment` is one of the 4 main compartments composing a fetal kidney.
We expect immune and endothelial cells to be healthy (non-cancerous) cells identified easily with high confidence.
We expect stroma and fetal nephron compartment to contain both normal and malignant cells.

@@

Here we check whether one (or more) `Seurat` cluster do correspond to one label.
This would allow us to be more confident in the annotation of immune or endothelial cells for example.

@@

#### Comparison of the  fetal reference with `SingleR` and `CellAssign`
